Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-15 docs/manuscript.Rmd
updated

@@

1. $H_0: cov(g, x) = 0; H_1: cov(g, x) \neq 0$; *the SNP associates with the exposure*
2. $H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0$; *the SNP associates with the outcome*
3. $H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0$; *the exposure associates with the outcome*
4. $H_0: cov(g, y - \hat{y}) \neq 0; cov(g, y - \hat{y}) = 0$; *the SNP is independent of the outcome when the outcome is adjusted for the exposure*

@@

### The influence of measurement error on mediation

### Using MR to infer the existence of causality

>>>>

We can model a system whereby some exposure $x$ has a causal influence $\beta_x$ on an outcome $y$ such that

In addition, the exposure is influenced by a SNP $g$ such that

Mediation-based analyses that test whether $x$ causally relates to $y$ impinge on evaluating if the influence of $g$ on $y$ can be accounted for by conditioning on $x$, such that

$$
cov(g, y - \hat{y}) = 0
$$

where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$

MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that

$$
\begin{aligned}
\hat{x} & = \hat{\beta}_g g \\
y & = \beta_{MR}\hat{x} + \epsilon_{MR}
\end{aligned}
$$

where $\beta_{MR} \neq 0$.

@@

1. $H_0: cov(g, x) = 0; H_1: cov(g, x) \neq 0$; *the SNP associates with the exposure*
2. $H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0$; *the SNP associates with the outcome*
3. $H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0$; *the exposure associates with the outcome*
4. $H_0: cov(g, y - \hat{y}) \neq 0; H_1: cov(g, y - \hat{y}) = 0$; *the SNP is independent of the outcome when the outcome is adjusted for the exposure*

@@

### 2017-02-15 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{01 February 2017}

@@

Measurement error will of course arise in other types of data too, for
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{12}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{8}, and its impact has been explored in the
epidemiological literature extensively\textsuperscript{13,14}. Given the
near-ubiquitous presence of measurement error in phenomic data it is
vital to understand its impact on the tools we use for causal inference.

@@

By utilising genetic instruments in different ways, mediation-based
analysis and MR models have properties that confer some advantages and
some disadvantages for reliable causal inference. An important advantage
of mediation-based analysis is that in order to infer the causal
direction the true underlying biology of the genetic instrument doesn't
necessarily need to be known as long as it associates with the putative
exposure. In the CIT framework (described fully in the Methods) for
example, the test statistic is different if you test for the exposure
causing the outcome or the outcome causing the exposure, allowing the
researcher to infer the direction of causality between two variables by
performing the test in both directions and choosing the model with the
strongest evidence. The CIT also has the valuable property of being able
to distinguish between several putative causal graphs that link the
traits with the SNP (Figure 1). Such is not the case for MR, where in
order to infer the direction of causality between two traits the
instrument must be known to influence the exposure primarily,
associating with the outcome only through the exposure.

@@

MR does however have some important advantages over the mediation-based
approaches. The mediation-based approaches require that the exposure,
outcome and instrumental variables are all measured in the same data,
whereas recent extensions to MR circumvent this requirement, allowing
causal inference to be drawn when exposure variables and outcome
variables are measured in different samples\textsuperscript{28}. This
has the crucial advantage of improving statistical power by allowing
analysis in much larger sample sizes, and dramatically expands the
breadth of possible phenotypic relationships that can be
evaluated\textsuperscript{26}. Additionally, when MR assumptions are
satisfied the method is substantially more robust to there being
measurement error in the exposure variable. Indeed instrumental variable
(IV) analysis was in part initially introduced as a correction for
measurement error in the exposure\textsuperscript{29}, whereas it has
been noted that both classic mediation-based
analyses\textsuperscript{13,14,30,31} and mediation-based methods that
use instrumental variables\textsuperscript{32,33} are prone to be
unreliable in its presence.

@@

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \(H_0: cov(g, x) = 0; H_1: cov(g, x) \neq 0\); \emph{the SNP
  associates with the exposure}
\item
  \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
  associates with the outcome}
\item
  \(H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0\); \emph{the exposure
  associates with the outcome}
\item
  \(H_0: cov(g, y - \hat{y}) \neq 0; cov(g, y - \hat{y}) = 0\);
  \emph{the SNP is independent of the outcome when the outcome is
  adjusted for the exposure}
\end{enumerate}

@@

\subsubsection{The influence of measurement error on
mediation}\label{the-influence-of-measurement-error-on-mediation}

\subsubsection{Using MR to infer the existence of
causality}\label{using-mr-to-infer-the-existence-of-causality}

>>>>

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{15 February 2017}

@@

Measurement error will of course arise in other types of data too. For
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{12}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{8}, and its impact has been explored in the
epidemiological literature extensively\textsuperscript{13,14}. Given the
near-ubiquitous presence of measurement error in phenomic data it is
vital to understand its impact on the tools we use for causal inference.

@@

By utilising genetic instruments in different ways, mediation-based
analysis and MR models have properties that confer some advantages and
some disadvantages for reliable causal inference. In the CIT framework
(described fully in the Methods) for example, the test statistic is
different if you test for the exposure causing the outcome or the
outcome causing the exposure, allowing the researcher to infer the
direction of causality between two variables by performing the test in
both directions and choosing the model with the strongest evidence. The
CIT also has the valuable property of being able to distinguish between
several putative causal graphs that link the traits with the SNP (Figure
1). Such is not the case for MR, where in order to infer the direction
of causality between two traits the instrument must be known to
influence the exposure primarily, associating with the outcome only
through the exposure.

@@

MR has some important advantages over the mediation-based approaches.
The mediation-based approaches require that the exposure, outcome and
instrumental variables are all measured in the same data, whereas recent
extensions to MR circumvent this requirement, allowing causal inference
to be drawn when exposure variables and outcome variables are measured
in different samples\textsuperscript{28}. This has the crucial advantage
of improving statistical power by allowing analysis in much larger
sample sizes, and dramatically expands the breadth of possible
phenotypic relationships that can be evaluated\textsuperscript{26}.
Additionally, when MR assumptions are satisfied the method is robust to
there being measurement error in the exposure variable. Indeed
instrumental variable (IV) analysis was in part initially introduced as
a correction for measurement error in the exposure\textsuperscript{29},
whereas it has been noted that both classic mediation-based
analyses\textsuperscript{13,14,30,31} and mediation-based methods that
use instrumental variables\textsuperscript{32,33} are prone to be
unreliable in its presence.

@@

\subsection{Model}\label{model}

We can model a system whereby some exposure \(x\) has a causal influence
\(\beta_x\) on an outcome \(y\) such that

In addition, the exposure is influenced by a SNP \(g\) such that

Mediation-based analyses that test whether \(x\) causally relates to
\(y\) impinge on evaluating if the influence of \(g\) on \(y\) can be
accounted for by conditioning on \(x\), such that

\[
cov(g, y - \hat{y}) = 0
\]

where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\)

MR analysis estimates the causal influence of \(x\) on \(y\) by using
the instrument as a proxy for \(x\), such that

\[
\begin{aligned}
\hat{x} & = \hat{\beta}_g g \\
y & = \beta_{MR}\hat{x} + \epsilon_{MR}
\end{aligned}
\]

where \(\beta_{MR} \neq 0\).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \(H_0: cov(g, x) = 0; H_1: cov(g, x) \neq 0\); \emph{the SNP
  associates with the exposure}
\item
  \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
  associates with the outcome}
\item
  \(H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0\); \emph{the exposure
  associates with the outcome}
\item
  \(H_0: cov(g, y - \hat{y}) \neq 0; H_1: cov(g, y - \hat{y}) = 0\);
  \emph{the SNP is independent of the outcome when the outcome is
  adjusted for the exposure}
\end{enumerate}

@@
